---
title: "R and dplyr"
author: "Pao"
date: "2024-04-18"
output: html_document
---

```{r}
library(dplyr)
```

* Use dplyr to create queries.
* DBI handles the backend associated to the database. In this case SQLite.

```{r}
connection <- DBI::dbConnect(RSQLite::SQLite(), 'data/penguins.db')

DBI::dbListTables(connection)
```
* Database connections are coordinated by the DBI package
* List tables with dbListTables()

```{r}
penguins <- dplyr::tbl(connection, 'penguins')
```

* Retrieve a table with tbl()
* It need a connection and the name of the table.

```{r}
penguins

penguins |> 
  dplyr::show_query()
```

* All dplyr functions (and call to the table) generates a `SELECT` statement.
* `show_query()` runs the SQL `EXPLAIN` command to show the SQL query.

## Filtering 

```{r}
penguins |> 
  filter(species == "Adelie")
```
* `filter()` generates a query using `WHERE <condition>` 
* dbplyr call are lazy. 
* The SQL query is only evaluated when is send to the database
* running the dplyr call only returns a preview of the result

```{r}
penguins |> 
  filter(species == "Adelie") |> 
  collect()
```

* The collect() function retrieve the complete result of the query (note that now the table has 152 rows).

## Selecting columns

```{r}
penguins |> 
  select(species, island, contains('bill'))
```

* `select()` modifies the `SELECT` statement.
* Selection helpers from dplyr (like `contains()`) works ok.

## Sorting

```{r}
penguins |> 
  select(species, body_mass_g) |> 
  arrange(body_mass_g)
```

* With `arrange()` you `ORDER BY` the data.
* There is a `desc()` function.

### Exersise

Find the lightest penguin in Dream Island by arranging the table accordingly. Only show the specie, island and body mass.

```{r}
penguins |> 
  select(species, island, body_mass_g) |> 
  filter(island == 'Dream') |> 
  arrange(desc(body_mass_g))
```
## Transforming columns 

```{r}
penguins |> 
  mutate(weight_kg = body_mass_g/1000)
```

* `mutate()` also modifies the `SELECT` statement.
  * Naming the new variable works as the `AS` statement. It is optional, but desirable. 
* If `select()` it not present the query will call all the columns.

## Agregating 

```{r}
penguins |> 
  group_by(species) |> 
  summarise(avg_body_mas = mean(body_mass_g))
```

* `summarise()` and `group_by()` work together to generate a `GROUP BY` clause.
* dplyr defaults to SQL to handle missing values unless you use the `na.rm` argument in aggregation functions.

### Excersise

* Calculate the bill ration (length/depth) for each penguin.
* Using the previous result, calculate the average ration for each specie.

## Creating a table

```{r}
library(DBI)

another_connection <- dbConnect(RSQLite::SQLite(), ':memory:')

table1 <- tibble(
  person = c('mik', 'mik', 'mik', 'po', 'po', 'tay'),
  job = c('calibrate', 'clean', 'complain', 'clean', 'complain', 'complain')
)

dbWriteTable(another_connection, "work", table1, row.names = FALSE, overwrite = TRUE)

work <- tbl(another_connection, "work")
```

* Writes, overwrites or appends a data frame to a database table with `dbWriteTable()` 
* `':memory:'` is a path that creates an in-memory database.

### Exersise

Create the job table, it should look like this:

|   name    | billable |
|-----------|----------|
| calibrate | 1.5      |
| clean     | 0.5      |

```{r}
table2 <- tibble(
  name = c('calibrate', 'clean'),
  billable = c(1.5, 0.5)
)

dbWriteTable(another_connection, "job", table2, row.names = FALSE, overwrite = TRUE)

job <- tbl(another_connection, "job")
```



Who doesn’t calibrate?

```{r}
work |> 
  filter(job != 'calibrate') |> 
  distinct(person)
```
* Similar to pure SQL, the result is wrong. Mik does calibrate. 
  * But using subqueries with dplyr is not that simple.

```{r eval=FALSE, include=FALSE}
work |> 
  filter(!job %in% (
    #subquery
    work |> 
      filter(job == 'calibrate') |> 
      distinct(person) |> 
      collect()
  )) |> 
  distinct(person)
```

* This does not work.

```{r}
work |> 
  filter(!person %in% sql(
    #subquery
    "(SELECT DISTINCT person FROM work
    where job = 'calibrate')"
  )) |> 
  distinct(person) 
```
* You can use literal SQL inside `sql()`. It returns an SQL object.
  * Useful when R code is not enough to write a query.

## Joining tables

```{r}
does_calibrate <- work |> 
  filter(job == 'calibrate')

work |> 
  anti_join(does_calibrate, by = join_by(person)) |> 
  distinct(person) |> 
  collect() 
```

* Joining tables with dplyr works similar to SQL. 
  * The `by` argument works as the `ON` statement. In this case we join by `person`.
* Use `left_join()`, `right_join()`, `inner_join()` or `full_join()`.
  
### Exersise 

How may hours each person works summing all jobs?

```{r}
work |> 
  left_join(job, by = join_by(job == name)) |> 
  group_by(person) |> 
  summarise(total_hours = sum(billable))
```

































































































